---
title: "Final project"
author: "Shayan Rahim"
date: "2023-05-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)



```





```{r}
library('leaflet')
library('dplyr')
library('tidyr')
library('ggplot2')
library('plotly')
library('tidyverse')
library('lubridate')
library("TSstudio")

```




```{r}
customer_shopping <- read.csv("customer_shopping_data.csv")
glimpse(customer_shopping)

```



### check for missing data
```{r}
sum(is.na(customer_shopping))
```


### check for duplicated data
```{r}
sum(duplicated(customer_shopping))
```


### check data types
```{r}
#check data type
sapply(customer_shopping, class)


```


### conver invoice_date data type from character to Date to be able to plot with the column
```{r}
customer_shopping[['invoice_date']] = as.Date(customer_shopping[['invoice_date']], format = '%d/%m/%Y')
sapply(customer_shopping, class)
```


```{r}
summary(customer_shopping)
```


### split data into 2 data sets for the years 2021 and 2022 by filtering
```{r}

customer_shopping <- customer_shopping %>% mutate(age_grp = ifelse(age < 10, "0's",
                              ifelse(age < 20, "10's",
                              ifelse(age < 30, "20's",
                              ifelse(age < 40, "30's",
                              ifelse(age < 50, "40's",
                              ifelse(age < 60, "50's", "60's")))))))

customer_shopping=customer_shopping %>% filter(invoice_date >= '2021-01-01' & invoice_date <= '2022-12-31')

customer_shopping_2021 = customer_shopping %>% filter(invoice_date >= '2021-01-01' & invoice_date <= '2021-12-31')

customer_shopping_2022 = customer_shopping %>% filter(invoice_date >= '2022-01-01' & invoice_date <= '2022-12-31')


summary(customer_shopping_2021)
```

#### categories
```{r}
unique(customer_shopping$category)
```

cat_2021 = customer_shopping_2021 %>% group_by(category, gender) %>%  summarise(average_spending_by_cat_gen = mean(price))

fspend = spending_by_cat %>% filter(gender=="Female") 


```{r}
spending_2021 = customer_shopping_2021 %>% group_by(invoice_date) %>%  summarise(sales = sum(price)) 

spending_2021 %>% ggplot() + geom_line(aes(x=invoice_date, y=sales))+ labs(title="2021")

spending_2022 = customer_shopping_2022 %>% group_by(invoice_date) %>%  summarise(sales = sum(price))

spending_2022 %>% ggplot() + geom_line(aes(x=invoice_date, y=sales))+ labs(title="2022")
```

```{r}
spend_cat_mall_2021 =  customer_shopping_2021 %>% select(category, shopping_mall, price)

ggplot(data=spend_cat_mall_2021)+geom_col(mapping=aes(x=shopping_mall, y=price, fill=category)) + labs(title="2021")

spend_cat_mall_2022 =  customer_shopping_2022 %>% select(category, shopping_mall, price)

ggplot(data=spend_cat_mall_2022)+geom_col(mapping=aes(x=shopping_mall, y=price, fill=category)) + labs(title="2022")


```


```{r}
cat = customer_shopping %>% group_by(category, invoice_date, price) %>% summarise(spending = sum(price)) 
cat %>% ggplot() + geom_col(aes(x=invoice_date, y=spending, fill=category))+facet_wrap(~category)
```

```{r}
age_gen = customer_shopping %>% group_by(category,gender) %>% summarise(n =n()) 
age_gen %>% ggplot() + geom_col(aes(x=category, y=n, fill=gender)) + theme(axis.text.x = element_text(angle = 90))


```
```{r}
age_cat = customer_shopping %>% group_by(category,age_grp) %>% summarise(n =n()) 
age_cat %>% ggplot() + geom_col(aes(x=age_grp, y=n, fill=age_grp)) + facet_wrap(~category)+theme(axis.text.x = element_text(angle = 90))

```

```{r}
gender_mall_cat = customer_shopping %>% group_by(category, shopping_mall,gender) %>% summarise(n =n()) 
gender_mall_cat %>% ggplot() + geom_col(aes(x=category, y=n, fill=gender))+facet_wrap(~shopping_mall) + theme(axis.text.x = element_text(angle = 90))

```


```{r}
gender_count= customer_shopping %>% group_by(gender) %>% summarise(n =n())
gender_count %>% ggplot() + geom_col(aes(x=gender, y=n, fill=gender))
```






```{r}
age_mall = customer_shopping %>% group_by(age_grp ,shopping_mall, category) %>% summarise(n =n()) 
age_mall %>% ggplot() + geom_col(aes(x=category, y=n, fill=age_grp))+facet_wrap(~shopping_mall) + theme(axis.text.x = element_text(angle = 90))

```
```{r}
age_spend = customer_shopping %>% group_by(age_grp ,price, gender) %>% summarise(spending = sum(price)) 
age_spend %>% ggplot() + geom_col(aes(x=age_grp, y=spending, fill=gender))

```
```{r}


```


