---
title: "Final project"
author: "Shayan Rahim"
date: "2023-05-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```





```{r}
library('dplyr')
library('tidyr')
library('ggplot2')
library('plotly')
library('tidyverse')
```




```{r}
customer_shopping <- read.csv("customer_shopping_data.csv")
glimpse(customer_shopping)

```



### check for missing data
```{r}
sum(is.na(customer_shopping))
```


### check for duplicated data
```{r}
sum(duplicated(customer_shopping))
```


### check data types
```{r}
sapply(customer_shopping, class)


```


### convert invoice_date data type from character to Date to be able to plot with the column
```{r}
customer_shopping[['invoice_date']] = as.Date(customer_shopping[['invoice_date']], format = '%d/%m/%Y')

sapply(customer_shopping, class)
```


```{r}
summary(customer_shopping)
```


### splitting data into 3 data sets 
```{r}

customer_shopping <- customer_shopping %>% mutate(age_grp = ifelse(age < 10, "0's",
                              ifelse(age < 20, "10's",
                              ifelse(age < 30, "20's",
                              ifelse(age < 40, "30's",
                              ifelse(age < 50, "40's",
                              ifelse(age < 60, "50's", "60's")))))))

customer_shopping = customer_shopping %>% filter(invoice_date >= '2021-01-01' & invoice_date <= '2022-12-31')

customer_shopping_2021 = customer_shopping %>% filter(invoice_date >= '2021-01-01' & invoice_date <= '2021-12-31')

customer_shopping_2022 = customer_shopping %>% filter(invoice_date >= '2022-01-01' & invoice_date <= '2022-12-31')


```

#### checking category column for the different categories
```{r}
unique(customer_shopping$category)
```

cat_2021 = customer_shopping_2021 %>% group_by(category, gender) %>%  summarise(average_spending_by_cat_gen = mean(price))

fspend = spending_by_cat %>% filter(gender=="Female") 


```{r}
spending_2021 = customer_shopping_2021 %>% group_by(invoice_date) %>%  summarise(sales = sum(price)) 

spending_2021 %>% ggplot() + geom_line(aes(x=invoice_date, y=sales))+ labs(title="2021") + labs(title="2021")+xlab("Time Period")+ylab('Sales')

```
```{r}
spending_2022 = customer_shopping_2022 %>% group_by(invoice_date) %>%  summarise(sales = sum(price))

spending_2022 %>% ggplot() + geom_line(aes(x=invoice_date, y=sales)) + labs(title="2022")+xlab("Time Period")+ylab('Sales')
```



### How much did people spend at each mall in 2021 and 2022 and how much by 
```{r}
spend_cat_mall_2021 =  customer_shopping_2021 %>% select(category, shopping_mall, price)

ggplot(data=spend_cat_mall_2021)+geom_col(mapping=aes(x=shopping_mall, y=price, fill=category)) + labs(title="2021") + theme(axis.text.x = element_text(angle = 90)) +xlab("Malls")+ylab('Spent')

spend_cat_mall_2022 =  customer_shopping_2022 %>% select(category, shopping_mall, price)

ggplot(data=spend_cat_mall_2022)+geom_col(mapping=aes(x=shopping_mall, y=price, fill=category)) + labs(title="2022")+ theme(axis.text.x = element_text(angle = 90)) +xlab("Malls")+ylab('Spent')


```



### Are there any patterns in each category? 
```{r}
cat_2021 = customer_shopping_2021 %>% group_by(category, invoice_date, price) %>% summarise(spending = sum(price)) 
cat_2021 %>% ggplot() + geom_col(aes(x=invoice_date, y=spending, fill=category))+facet_wrap(~category) + labs(title="2021")+ theme(axis.text.x = element_text(angle = 90)) +xlab("Timeline")+ylab('Spending')

cat_2022 = customer_shopping_2022 %>% group_by(category, invoice_date, price) %>% summarise(spending = sum(price)) 
cat_2022 %>% ggplot() + geom_col(aes(x=invoice_date, y=spending, fill=category))+facet_wrap(~category)+ labs(title="2022")+ theme(axis.text.x = element_text(angle = 90)) +xlab("Timeline")+ylab('Spending')


```



### Let's take a look at clothing category for 2021. See any pattern?
```{r}
clothing_2021 = customer_shopping_2021 %>% group_by(category, invoice_date, price) %>% filter(category=="Clothing") %>%summarise(spending = sum(price)) 
clothing_2021 %>% ggplot() + geom_line(aes(x=invoice_date, y=spending, color=category))+ labs(title="2021")+ theme(axis.text.x = element_text(angle = 90)) +xlab("Timeline")+ylab('Spending')

```



### Let's visualize the number of purchases by gender and age group
```{r}
gender_count= customer_shopping %>% group_by(gender, age_grp) %>% summarise(n =n())
gender_count %>% ggplot() + geom_col(aes(x=gender, y=n, fill=age_grp)) +labs(title=" Number of Purchases by Genders and Age Groups")+xlab("Gender")+ylab('Number of Items')

```



###  Is there a relationship between gender and category?
```{r}
age_gen = customer_shopping %>% group_by(category,gender) %>% summarise(n =n()) 
age_gen %>% ggplot() + geom_col(aes(x=category, y=n, fill=gender)) + theme(axis.text.x = element_text(angle = 90))+labs(title="Purchases by Gender")+xlab("Category")+ylab('Number of Purchases')


```



### How about different age groups and categories
```{r}
age_cat = customer_shopping %>% group_by(category,age_grp) %>% summarise(n =n()) 
age_cat %>% ggplot() + geom_col(aes(x=age_grp, y=n, fill=age_grp)) + facet_wrap(~category)+theme(axis.text.x = element_text(angle = 90)) + labs(title="Purchases by Age Group")+xlab("Category")+ylab('Number of Purchases')

```



### Do certain demographics of people go to different malls for different categories of shopping
```{r}
gender_mall_cat = customer_shopping %>% group_by(category, shopping_mall,gender) %>% summarise(n =n()) 
gender_mall_cat %>% ggplot() + geom_col(aes(x=category, y=n, fill=gender))+facet_wrap(~shopping_mall) + theme(axis.text.x = element_text(angle = 90)) + labs(title="Mall Demographics by Gender")+xlab("Category")+ylab('Number of Purchases')

```


### Let's take age groups into account as well
```{r}
age_mall = customer_shopping %>% group_by(age_grp ,shopping_mall, category) %>% summarise(n =n()) 
age_mall %>% ggplot() + geom_col(aes(x=category, y=n, fill=age_grp))+facet_wrap(~shopping_mall) + theme(axis.text.x = element_text(angle = 90)) + labs(title="Mall Demographics by Gender and Age Groups")+xlab("Category")+ylab('Number of Purchases')

```




### What about spending by age group?
```{r}
age_spend = customer_shopping %>% group_by(age_grp ,price, gender) %>% summarise(spending = sum(price)) 
age_spend %>% ggplot() + geom_col(aes(x=age_grp, y=spending, fill=gender)) + labs(title="Spending vs Age Groups")+xlab("Age Groups")+ylab('Spending')

```




